---
title: Bioenergetic crisis in Intensive Care Unit Acquired Weakness gene signatures were associated with sepsis-related mortality
author: Seibi Kobara
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float: yes
    toc_collapse: no
---

```{r include=FALSE}
library(limma)
library(GEOquery)
library(pheatmap)
library(topGO)
library(AnnotationDbi)
library(illuminaHumanv4.db)
library(clusterProfiler)
library(ggplot2)
library(tidyverse)
library(magrittr)
library(egg)
library(annotate)
library(pathview)
library(ggpubr)
```

# Import file
```{r message=FALSE}
gse <- getGEO("GSE54514")[[1]]
```

# Inspect clinical variables
```{r message=FALSE}
# Sample information
sampleInfo <- pData(gse)

# Select variables you want to store
sampleInfo <- dplyr::select(sampleInfo,
                     "age (years):ch1",
                     "disease status:ch1",
                     "gender:ch1",
                     "group_day:ch1",
                     "group_id:ch1",
                     "severity (apacheii):ch1",
                     "site of infection:ch1")

# Rename to more convenient column names
sampleInfo <- dplyr::rename(sampleInfo,
                     age="age (years):ch1",
                     disease_status="disease status:ch1",
                     gender="gender:ch1",
                     group_day="group_day:ch1",
                     group_ID="group_id:ch1",
                     APACHE2= "severity (apacheii):ch1",
                     site_of_infection="site of infection:ch1")

# Categorization of sample information (age, APACHE II)
sampleInfo$age<-as.numeric(sampleInfo$age)
sampleInfo$APACHE2<-as.numeric(sampleInfo$APACHE2)
sampleInfo$Age<-factor(ifelse(sampleInfo$age<40,1,
                                       ifelse(sampleInfo$age>=40 & sampleInfo$age<60,2,
                                              ifelse(sampleInfo$age>=60 & sampleInfo$age<80,3,4))),
                         levels = c(1:4),
                         labels = c("<40","40-60","60-80","<80"))
sampleInfo$APACHE_II<-factor(ifelse(is.na(sampleInfo$APACHE2),NA,
                            ifelse(sampleInfo$APACHE2<15,1,
                            ifelse(sampleInfo$APACHE2>=15 &sampleInfo$APACHE2<20,2,
                            ifelse(sampleInfo$APACHE2>=20 &sampleInfo$APACHE2<25,3,4)))),
                         levels = c(1:4),
                         labels = c("<15","15-20","20-25","<25"))

sampleInfo$Mortality <- ifelse(sampleInfo$disease_status =="sepsis nonsurvivor","Non-Survivor",
                               ifelse(sampleInfo$disease_status =="sepsis survivor","Survivor","healthy"))


# ID, group
sampleInfo<-tidyr::separate(sampleInfo,col="group_ID",into=c("group","ID"),sep="_")

# total population (lung and bacteremia)
# only day 1
sampleInfo1 <- sampleInfo %>% filter(group_day %in% c("NS_D1","S_D1")) %>%
    filter(site_of_infection %in% c("blood","lung","UT")) 
saveRDS(sampleInfo1,"sampleInfo.rds")
```

# Preprocess
```{r}
# Preprocess
# calculate median expression level
ex<-exprs(gse)
dim(ex)
# 24840   163
cutoff <- median(ex)
isexpressed<- ex > cutoff

# expressed more than 3 samples
keep <- rowSums(isexpressed) >2
table(keep)
# FALSE  TRUE 
#  863 23977 

# available genes
ex1<- ex[keep,]
dim(ex1 )
# 23977   163

# annotation
anno<- AnnotationDbi::select(illuminaHumanv4.db, keytype = "PROBEID", 
                             keys =rownames(ex1), 
                            columns = c("ENTREZID") )
ex1<-as.data.frame(ex1)
ex1 %<>% mutate(affy=rownames(ex1))
# merge
ex2<- merge(ex1, anno, all.x=T, by.x="affy",by.y="PROBEID")

# drop affy
ex3<-ex2[,-c(1)]

# Remove NA in entrez ID
table(is.na(ex3$ENTREZID))
# FALSE  TRUE 
# 19658  5264 

ex4<-ex3[!is.na(ex3$ENTREZID),]
dim(ex4)
# 19658   164

# Duplicated 
table(duplicated(ex4$ENTREZID)) 
# FALSE  TRUE 
# 13972  5686

# Remove duplicate by taking median
# only entrezID and expression
ex5 <- ex4 %>% group_by(ENTREZID) %>% summarise_all(median)
dim(ex5)
saveRDS(ex5,"expression.rds")
# 13972   164
```

# ICUAW associated genes from GSE78929
```{r}
# load data
AWgenelist<-readRDS("ICUAW_associated_genes.rds")
dim(AWgenelist)

# annotation: illumina to entrez
AW_list<- AnnotationDbi::select(illuminaHumanv4.db, keytype = "PROBEID", 
                             keys =AWgenelist$illumina_ID, 
                            columns = c("ENTREZID") )

# remove unavailable for entrez
AW_list<-AW_list[!is.na(AW_list$ENTREZID),] 
dim(AW_list)
# 675 2
# remove duplicated
AW_list2 <- AW_list[!duplicated(AW_list$ENTREZID),]
dim(AW_list2)
# 608 2
```

```{r}
#expression set for ICUAW
exp<-ex5 %>% filter(ENTREZID %in% AW_list$ENTREZID) 
expression <- exp %>% dplyr::select(rownames(sampleInfo))
expression %<>% mutate(ENTREZID=exp$ENTREZID)
saveRDS(expression,"ICUAWexpression.rds")
```

# Primary analysis 
# Differential gene expression analysis between non-survivors and survivors among patients with primary bacteremia- and respiratory-triggered sepsis
```{r}
# load
sample<-readRDS("sampleInfo.rds")          # 21 samples  11 variables
expression<-readRDS("ICUAWexpression.rds") # 526 36 (35 + entrez)

# expression
expression<- expression %>% dplyr::select(ENTREZID,rownames(sample))
dim(expression)
# 526 22 (21 + entrez)

# Model matrix
design <- model.matrix(~0+sample$Mortality + sample$age)
# Rename column name
colnames(design) <- c("non_surv","surv","age")

# Fitting
fit <- lmFit(expression[,-c(1)], design)

# Make contrast set
contrasts<- makeContrasts(non_surv - surv,levels=design)

# Fit
fit2 <- contrasts.fit(fit, contrasts)

# Bayes 
fit2 <- eBayes(fit2)

# annotation
anno<- AnnotationDbi::select(illuminaHumanv4.db, keytype = "ENTREZID", 
                             keys =as.character(expression$ENTREZID), 
                            columns = c("SYMBOL") )

fit2$genes <- anno

# Toptable
toptable <- topTable(fit2,coef = 1, number=Inf)
toptable %<>% mutate(direction=ifelse(logFC>0, "up","down"))
```
## GO enrichment analysis
```{r}
# EnrichGO for all
sig_gene<- toptable %>% filter(adj.P.Val<0.05 & abs(logFC)>0.5)

direction<-c("up","down")
ontology<-c("BP","CC","MF")
ego_all<-data.frame()
for (i in direction){
  df<-sig_gene %>% filter(direction==i)
  for(j in ontology){
  ego <- enrichGO(gene = df[[1]],
                keyType = "ENTREZID",
                OrgDb = org.Hs.eg.db, 
                pAdjustMethod = "BH",
                pvalueCutoff = 0.05,
                qvalueCutoff = 0.05,
                minGSSize = 10,
                maxGSSize = 500,
                ont = j,
                readable = TRUE)
  ego<- data.frame(ego)
  ego <- ego %>% mutate(direction=i)
  ego <- ego %>% mutate(ontology=j)
  ego_all<-rbind(ego_all,ego)
  }}
```

## KEGG pathway enrichment analysis
```{r}
# for all
sig_gene<- toptable %>% filter(adj.P.Val<0.05 & abs(logFC)>0.5)

KEGG_all<-data.frame()
for (i in direction){
  df<-sig_gene %>% filter(direction==i)
  k <- enrichKEGG(gene = df[[1]], organism = 'hsa',pvalueCutoff = 0.05,)
  k <-as.data.frame(k)
  k<- k %>% mutate(direction=i)
  KEGG_all<-rbind(KEGG_all,k)
}
```

# Secondary analysis 
# Differential gene expression analysis among respiratory-triggered sepsis

```{r}
soi<-c("blood","lung") # soi: site of infection
toptable_soi <- data.frame()
for(i in soi){
  # Model matrix
  sample_select<-sample %>% filter(site_of_infection==i)
  exp<- expression %>% dplyr::select(rownames(sample_select))
  design <- model.matrix(~0+sample_select$Mortality + sample_select$age)
  # Rename column name
  colnames(design) <- c("non_surv","surv","age")
  # Fitting
  fit <- lmFit(exp, design)
  # Make contrast set
  contrasts<- makeContrasts(non_surv - surv, levels=design)
  # Fit
  fit2 <- contrasts.fit(fit, contrasts)
  # Bayes 
  fit2 <- eBayes(fit2)
  # annotation
  anno<- AnnotationDbi::select(illuminaHumanv4.db, keytype = "ENTREZID", 
                             keys =as.character(expression$ENTREZID), 
                            columns = c("SYMBOL") )
  fit2$genes <- anno
  # Toptable
  toptable <- topTable(fit2,coef = 1, number=Inf)
  toptable %<>% mutate(direction=ifelse(logFC>0, "up","down"))
  toptable %<>% mutate(soi=i)
  toptable_soi <- rbind(toptable_soi,toptable)
}
```

## GO enrichment analysis

```{r}
# EnrichGO for soi
sig_gene<- toptable_soi %>% filter(adj.P.Val<0.05 & abs(logFC)>0.5)

direction<-c("up","down")
ontology<-c("BP","CC","MF")
soi<- c("lung","UT")
ego_soi<-data.frame()
for(k in soi){
  df0<-sig_gene %>% filter(soi==k)
  for (i in direction){
    df<-df0 %>% filter(direction==i)
    for(j in ontology){
      ego <- enrichGO(gene = df[[1]],
                keyType = "ENTREZID",
                OrgDb = org.Hs.eg.db, 
                pAdjustMethod = "BH",
                pvalueCutoff = 0.05,
                qvalueCutoff = 0.05,
                minGSSize = 10,
                maxGSSize = 500,
                ont = j,
                readable = TRUE)
      ego<- data.frame(ego)
      ego %<>% mutate(soi=k)
      ego %<>% mutate(direction=i)
      ego %<>% mutate(ontology=j)
      ego_soi<-rbind(ego_soi,ego)
}}}
```

## KEGG pathway enrichment analysis
```{r}
# for soi
sig_gene<- toptable_soi %>% filter(adj.P.Val<0.05 & abs(logFC)>0.5)
soi<- c("lung","UT")
KEGG_soi<-data.frame()
for(k in soi){
  df0<- sig_gene %>% filter(soi==k)
  for (i in direction){
    df<- df0 %>% filter(direction==i)
    kegg <- enrichKEGG(gene = df[[1]], organism = 'hsa',pvalueCutoff = 0.05,)
    kegg <- data.frame(kegg)
    kegg %<>% mutate(direction=i)
    kegg %<>% mutate(soi=k)
    KEGG_soi<-rbind(KEGG_soi,kegg)
}}
```
